\texttt{vqtl}: An \texttt{R} package for Mean-Variance QTL Mapping

### 2018-07-10 revision/response.tex

\usepackage{color}
\usepackage[usenames,dvipsnames]{xcolor}

\newcommand{\alert}[1]{\textcolor{red}{#1}}
\usepackage[left,running,mathlines]{lineno}
\renewcommand\linenumberfont{\color{gray}\normalfont\tiny\sffamily}
\usepackage{listings}

\newcommand{\auassign}[2]{\mbox{}\marginpar{\footnotesize\textbf{#1}\newline \textcolor{red}{#2}}}

\lhead{Manuscript ID: G3/2018/200159 \\ Title: ``\texttt{vqtl}: An \texttt{R} package for Mean-Variance QTL Mapping''}
\chead{\bfseries Response to Reviewer Comments \\ \mbox{}\\ }
\rhead{\today\\William Valdar}
\title{BVH: Response To Editors/Reviewers}

@@

\EditorQuestion{
  Now that you have all three manuscripts potentially going forward for publication in G3, they can be more tightly linked.
}
\Response{
  We have updated the manuscripts such that each one refers to the other two in the introduction and discussion.
}

\EditorQuestion{
   For instance, one of the other papers could provide the benchmark data that is required for [the software] manuscript.
   So the software paper can include those files and refer to the other manuscript for details.
}
\Response{
  We have incorporated this recommendation, using the dataset from Kumar et al. 2013 that was used in one of the case studies in the manuscript titled ``Mean-Variance QTL Mapping Identifies Novel QTL for Circadian Activity and Exploratory Behavior in Mice''.
  The time required to analyze this dataset, according to number of computer cores available and number of permutations desired, is described in the ``Performance Benchmarks'' section of the software paper and the relevant files are included in its Zenodo repository.
}

\subsection*{Issues Specific to this Manuscript}

\EditorQuestion{
  As suggested by Reviewer 1 it will be very useful to have a benchmark study with times/memory requirements as a function of number of markers and sample size.
  This study will give those interested on using the package an understanding of how this package would scale.
  It will be important to include in the discussion also elements related to scalability: is the package suitable for very large-p/large-n? 
}
\Response{
  We have added this information in the ``Performance Benchmarks'' section. 
  See detailed response to Reviewer 1 comment 3. 
  Note that this has lengthened the paper somewhat, but we can move plots to supplemental if that would be preferred.
}

@@

% Sørensen P, de los Campos G, Morgante F, Mackay TFC, Sorensen D. Genetic Control of Environmental Variation of Two Quantitative Traits of Drosophila melanogaster Revealed by Whole-Genome Sequencing. Genetics. 2015;201(2):487-497. doi:10.1534/genetics.115.180273.

\section*{Reviewer 1 Comments}

\ReviewerQuestion{
  The authors mention ``in almost all traits of interest in human health and disease''.
  The detection of QTL has also been widely used in plant and animal populations.
}
\Response{
  The reviewer is quite correct and we have edited the introduction to reflect the fact that QTL mapping has been applied productively in plant and animal research.
}

\ReviewerQuestion{
  The authors used a set of simulated data with 3 chromosomes and 11 equally spaced markers.
  I would like to see an example with a larger genotype data set.
}
\Response{
    Our use of an unrealistically-small genome is to most clearly illustrate the standard workflow.
%    But we are encouraged by the reviewer's enthusiasm to see the package applied to real datasets.
    Since this manuscript has been submitted as part of a three-paper companionship where each of the other papers contains analyses of three full sets of real data using the R package, we feel that using a reduced dataset to demonstrate the package functions is reasonable in this instance.
    Nonetheless, we note that a full, real dataset is used in the newly added performance benchmark, as described below.
}

\ReviewerQuestion{
  The CPU time is indicated for the data set of the example.
  I would like to have the CPU time information in a more realistic data set.
  Perhaps authors can use a free data set from the web.
}
\Response{
  Thank you for this recommendation.
  The authors agree that including this information will make the paper more useful to geneticists who are considering using the package --- we have added two sets of benchmarks.
}
\Response{
  First, we demonstrate the performance of the package according to the number of computer cores used and the number of permutations conducted.
  For this set of benchmarks, we used the intercross from Kumar et. al 2013, which we reanalyzed in \CortyReanalysisPaper.
  This intercross population has 244 mice and was genotyped at 93 markers.
  After applying the hidden Markov model in \texttt{R/qtl} with \texttt{step.size} of 2, the number of loci to be tested was 582.
  As recommended by the editor, this example helps tie together the companions, because the results of this analysis are presented in the manuscript titled ``Mean-Variance QTL Mapping Identifies Novel QTL for Circadian Activity and Exploratory Behavior in Mice''.
  This addition can be found on page 4.
}
\Response{
  Second, we demonstrate the performance of the package according to the size of the mapping population and the number of genetic markers, holding constant the number of permutations at 1000 and the number of cores at 32.
  For this set of benchmarks, we used crosses simulated with \texttt{R/qtl} so we could control the population size and number of genetic markers.
  This addition can be found on page 4.
}

### 2018-07-10 Package_vqt_Manuscript.tex
edits

\author[$\ast$,1]{Robert W. Corty}
\author[$\ast, 1$]{William Valdar}

\affil[$\ast$]{University of North Carolina at Chapel Hill, Department of Genetics}

\keywords{QTL mapping, vQTL, mvQTL, variance heterogeneity, DGLM, R}

@@

QTL mapping studies in experimental crosses have provided important insights on nearly every trait of interest in human health, as well as livestock and crop improvement programs.
Advances in model organism genotyping \citep{Williams1990} and phenotyping \citep{Yang2014a} as well as in statistical methods \citep{Lander1989a,Martinez1992} and software tools (\eg, \citealt{Broman2003,Mulligan2017}) have supported these discoveries.

It has long been recognized that the extent of residual variation is itself a heritable trait \citep{Falconer1965,Lynch1998}.
This phenomenon has been described theoretically \citep{Hill2004-uo,Hill2010} and demonstrated in inbred model organisms \citep{Sorensen2015,Ayroles2015} and crops \citep{Yang2012-aw,Forsberg2015} and exploited in livestock improvement efforts \citep{Mulder2008,Ibanez-Escriche2008-ie}.

However, traditional QTL mappinng analyses have focused on discovering ``mean QTL'', regions of the genome where allelic variation drives heterogeneity of phenotype mean, while assuming that the residual variance, that is, the intrinsic stability or noisiness of the phenotype, is identical for each every individual in the mapping population.
Recent work, however, has challenged the assumption of homogeneous variance and sought to identify an additional type of QTL --- those that
%, by driving heterogeneity of epistatic or environmental variance, 
influence the extent of residual variance, sometimes termed ``variance QTL'' (vQTL) \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014}.

Although detection of vQTL has started to enter the mainstream of genetic analysis \citep{Yang2012,Hulse2013,Ayroles2015,Wei2016-lt,Wang2017,Wei2017-tt}, the statistical methods used for this purpose remain heterogeneous.
% and sometimes idiosyncratic, with connections to methods for mapping mean QTL often lacking or unclear.
% a clear or intuitive connection to methods used for detecting mean QTL.
%there remains great heterogeneity among the statistical methods used to conduct these analyses.
We developed a standardized method for QTL mapping that can detect mQTL, vQTL, and a generalization of the two that we term ``mvQTL''.
This approach, which we term ``mean-variance QTL mapping'', can be applied to intercross and backcross designs.
In two companion articles, we characterize this method and competitors in the setting where a background factor drives variance heterogeneity (\CortyMethodsPaper), and use it to discover new QTL from two existing data resources (\CortyReanalysisPaper)

Here, we provide a practical guide to using the \texttt{R} package \texttt{vqtl}, which implements mean-variance QTL mapping.
First, to generate illustrative data, we simulate an F2 intercross and four phenotypes: one phenotype determined entirely by random noise, and one with each of the three kinds of QTL.
On each phenotype we then conduct a genome scan using standard approximations to interval mapping \citep{Lander1989a,Martinez1992}, and mean-variance QTL mapping, which includes a test for mQTL, a test for vQTL, and a test for mvQTL.
The association statistics of all four tests are then initially plotted in LOD score units, with drawbacks of this plotting unit discussed.
Permutation scans are used to determine empirically adjusted $p$-values, and plotting in these units is shown to to make the results of the four tests more comparable.
Last, we describe plots to communicate the effects that led to the detection of a QTL, and use the bootstrap to estimate its confidence interval.

In two companion articles, we apply mean-variance QTL mapping with package \texttt{vqtl} to discover new QTL.
In the first, we describe a novel vQTL for exploratory behavior and a novel mQTL for a circadian behavior trait, both in mouse F2 intercrosses retrieved from the Mouse Phenome Database (\CortyReanalysisPaper) \citep{Bogue2015}.
The dataset from one of these intercrosses is used to benchmark the performance of package \texttt{vqtl}, as is described in the ``Performance Benchmarks'' section.
In the second, we describe a unique strength of mean-variance QTL mapping, its ability to accommodate variance heterogeneity arising from background factors, such as sex, batch, and housing, and demonstrate the value of such accommodation with both simulation studies and a report of a novel QTL (\CortyMethodsPaper).

% Said another way, it is typically assumed that nothing -- neither environmental factors nor genetic factors -- influences the residual variance of the trait.

% In fact, gene-gene interactions, gene-environment interactions, genetic factors that influence microenvironmental sensitivity, and environmental factors can influence residual variance.

% This approach allows for QTL mapping in the presence of both genetic and non-genetic effects on phenotype mean and variance.
% We compare its behavior to other QTL mapping approaches, and illustrate the discovery of a new QTL from an existing study.

% overview of the software or resource
% In support of other researchers who may be interested in mvQTL mapping, we created \texttt{R} package \texttt{vqtl}, which provides functions for conducting genome scans, assessing the statistical significance of results, and visualizing and interpreting significant findings.

% describe how researchers can access it.
% \texttt{R} package \texttt{vqtl} uses the same \texttt{cross} data structure as the popular \texttt{qtl} package and is available on \texttt{CRAN}, so it is easy to get started.
% Here, we demonstrate typical use of the \texttt{vqtl} package to analyze four simulated traits and discover three QTL.
% a QTL that influence phenotype mean (an mQTL), a QTL that influences phenotype variance (vQTL) and a QTL that influences a combination of mean and variance (an mvQTL).
% The code used to simulate the phenotypes, calculate the statistics, estimate their significance, and visualize significant results is available at \texttt{github.com/rcorty}.

@@

To illustrate the use of the \texttt{vqtl} package, we first simulated an example F2 intercross using the popular \texttt{R/qtl} package \citep{Broman2003}, on which \texttt{vqtl} is based.
This cross consisted of 200 male and 200 female F2 offspring, with 3 chromosomes of length 100 cM, each tagged by 11 equally-spaced markers and estimated genotype probabilities at 2cM intervals with \texttt{R/qtl}'s hidden Markov model. We then generated four phenotypes:
\begin{enumerate}
	\item \texttt{phenotype1} consists only of random noise and will serve as an example of negative results for all tests.
	\item \texttt{phenotype2} has an mQTL that explains 4\% of phenotype variance at the center of chromosome one.
	\item \texttt{phenotype3} has a vQTL at the center of chromosome two.
        This vQTL acts additively on the log standard deviation scale, and results in residual standard deviation of [0.8, 1, 1.25] for the three genotype groups.
	\item \texttt{phenotype4} has an mvQTL at the center of chromosome three.
        This mvQTL has a mean effect that explains 2.7\% of phenotype variance and a variance effect that acts additively on the standard deviation scale, resulting in residual standard deviation of [0.85, 1, 1.17] for the three genotype groups.
\end{enumerate}

@@

The $p$-values described above, however, are nominal: they do not take into account multiple testing across the genome.
\RC{
    They also rely on asymptotic theory that assumes the underlying phenotype being residually normal; this may not always be the case and when violated will lead to inflated significance.
}{
    Could change this to read ``They also rely on the alternative model being exactly true; severe model misspecification is severe this can lead to inflated statistics.''
    Would be more accurate, but I don't know the rules with regards to things the reviewer didn't mention.
}

More robust $p$-values that are corrected for genomewide significance via control of the family-wise error rate (FWER) can be obtained empirically, through a permutation procedure described below.

@@

Last, it is important to assess the genetic precision of a discovered QTL for bioinformatic follow-up.
The function \texttt{scanonevar.boot} implements the non-parametric bootstrap \citep{Visscher1996}.
This function takes, as arguments, a \texttt{scanonevar} object, the type of QTL detected, the name of the chromosome containing the QTL, and \texttt{num.resamples}, the number of bootstrap resamplings desired.
As with \texttt{scanonevar.perm}, the \texttt{n.cores} argument can be used to spread the bootstraps over many computational cores and defaults to the number of cores available minus two, and bootstraps can be run on separate computers and combined with \texttt{c} to increase the precision of the estimate of the confidence interval.

@@

For the first benchmark, we ran \texttt{scanonevar.perm} on the data from \citet{Kumar2013} and \CortyReanalysisPaper, which contains 244 individuals and 582 loci, varying the number of permutations desired and the number of computer cores used.
For a given number of cores, there is a linear relationship between number of permutations conducted and time required \autoref{fig:benchmark_kumar}.
The slope the the line indicates time required per permutation and is dependent on the number of cores, ranging from $\approx$ 6.3 seconds per permutation with 4 cores to $\approx$ 1.2 second per permutation with 32 cores.

For the second benchmark, we ran \texttt{scanonevar.perm} on simulated data, always conducting 1000 permutations and using 32 cores, but varying the number of individuals in the mapping population and the number of markers in the genome.
For a given population size, there is a slightly supra-linear relationship between number of markers and time required \autoref{fig:benchmark_sim_cross}, which reflects a linear increase in the time taken to conduct the permuted genome scans, plus an increase in the time taken for ``bookkeeping'' tasks like organizing and reshaping genetic data.
The slope of the line indicates the time required per locus and is dependent on the population size, ranging from $\approx$ 1.4 seconds per locus with a population of size 100 to $\approx$ 3.3 seconds per locus with a population of size 800.

Based on these benchmarks, it is clear that the workflow presented here is practical for QTL mapping F2 intercross and similar populations on modern, multi-core scientific computers.
Populations with many recombinations, where dense genotyping arrays that interrogate $\textgreater$ 10,000 loci are relevant, could not be practically analyzed with package \texttt{vqtl} in this way.
However, both statistical and computational steps could be taken to make such a study feasible.
Statistically, techniques could be used that allow for large-scale analysis without permutation testing \citep{Efron2004}.
Computationally, the software could be changed to run on a computer cluster, rather than on a single computer \citep{Jette2003a,Marchand2017}.

@@

\section{Acknowledgments}
This work was primarily funded by a National Institutes of General Medical Sciences (NIGMS) grant to WV (R01-GM104125) and a National Institutes of Mental Health grant to RWC (F30-MH108265).
RWC received additional support from NIGMS grant T32-GM067553 and National Library of Medicine grant T32-LM012420.

>>>>

\author[$\ast,\dagger$]{Robert W. Corty}
\author[$\ast,\ddagger,1$]{William Valdar}

\affil[$\ast$]{Department of Genetics}
\affil[$\dagger$]{Bioinformatics and Computational Biology Curriculum}
\affil[$\ddagger$]{and Lineberger Comprehensive Cancer Center, University of North Carolina,
    Chapel Hill, NC}

\keywords{heteroskedastic regression, mvQTL, variance heterogeneity, DGLM}

@@

Traditional quantitative trait locus (QTL) analyses have focused on discovering ``mean QTL'', regions of the genome where allelic variation drives heterogeneity of phenotype mean, while assuming that the residual variance, that is, the intrinsic stability or noisiness of the phenotype, is identical for every individual in the mapping population.
Recent work, however, has challenged the assumption of homogeneous residual variance and sought to identify an additional type of QTL---those that influence the extent of residual variance, sometimes termed ``variance QTL'' (vQTL) \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014}.

Although detection of vQTL has started to enter the mainstream of genetic analysis \citep{Yang2012,Hulse2013,Ayroles2015,Wei2016-lt,Wang2017,Wei2017-tt}, statistical tools for this purpose remain heterogeneous.
We have developed a standardized method for QTL mapping in experimental crosses that simultanously model mean and variance effects in order to detect mQTL, vQTL and a generalization of the two that we term ``mvQTL''.
Our approach, which we term ``mean-variance QTL mapping'', can currently be applied to F2 intercross and backcross designs.
In two companion articles, we characterize this method and competitors in the setting where a background factor drives variance heterogeneity (\CortyMethodsPaper), and use it to discover new QTL from two existing data resources (\CortyReanalysisPaper).

Here, we provide a practical guide to the approach using its associated \texttt{R} package \texttt{vqtl}, which implements mean-variance QTL mapping and is inter-operable with the established \texttt{qtl} package \citep{Broman2003}.
First, to generate illustrative data, we simulate an F2 intercross and four phenotypes: one phenotype determined entirely by random noise, and one with each of the three kinds of QTL.
On each phenotype we then conduct a genome scan using standard approximations to interval mapping \citep{Lander1989a,Martinez1992}, and mean-variance QTL mapping, which includes a test for mQTL, a test for vQTL, and a test for mvQTL.
The association statistics of all four tests are initially plotted in LOD score units, with drawbacks of this plotting unit discussed; then permutation scans are used to determine empirically-adjusted $p$-values, and plotting in these units is shown to to make the results of the four tests more easily comparable.
Plots are then described that communicate effects that led to the QTL's detection, and the bootstrap is used to estimate its confidence interval.
Last, we benchmark performance, using one of the datasets examined in \CortyReanalysisPaper to report how computation time varies with marker density and number of permutations.

%In two companion articles, we apply mean-variance QTL mapping with package \texttt{vqtl} to discover new QTL.
%~In the first, we describe a novel vQTL for exploratory behavior and a novel mQTL for a circadian behavior trait, both in mouse F2 intercrosses retrieved from the Mouse Phenome Database (\CortyReanalysisPaper) \citep{Bogue2015}.
%In the second, we describe a unique strength of mean-variance QTL mapping, its ability to accommodate variance heterogeneity arising from background factors, such as sex, batch, and housing, and demonstrate the value of such accommodation with both simulation studies and a report of a novel QTL (\CortyMethodsPaper).

% \section*{Old intro material -- TO DELETE OR REPURPOSE}

% Quantitative Trait Locus (QTL) mapping studies in experimental crosses have provided important insights on nearly every trait of interest in human health, as well as livestock and crop improvement programs.
% Advances in model organism genotyping \citep{Williams1990} and phenotyping \citep{Yang2014a} as well as in statistical methods \citep{Lander1989a,Martinez1992} and software tools (\eg, \citealt{Broman2003,Mulligan2017}) have supported these discoveries.

% It has long been recognized that the extent of residual variation is itself a heritable trait \citep{Falconer1965,Lynch1998}.
% This phenomenon has been described theoretically \citep{Hill2004-uo,Hill2010} and demonstrated in inbred model organisms \citep{Sorensen2015,Ayroles2015} and crops \citep{Yang2012-aw,Forsberg2015} and exploited in livestock improvement efforts \citep{Mulder2008,Ibanez-Escriche2008-ie}.

@@

To illustrate the use of the \texttt{vqtl} package, we first simulated an example F2 intercross using \texttt{R/qtl} package \citep{Broman2003}, on which \texttt{vqtl} is based.
This cross consisted of 200 male and 200 female F2 offspring, with 3 chromosomes of length 100 cM, each tagged by 11 equally-spaced markers and estimated genotype probabilities at 2cM intervals with \texttt{R/qtl}'s hidden Markov model. We then generated four phenotypes:
\begin{enumerate}
	\item \texttt{phenotype1} consists only of random noise and will serve as an example of negative results for all tests.
	\item \texttt{phenotype2} has an mQTL that explains 4\% of phenotype variance at the center of chromosome one.
	\item \texttt{phenotype3} has a vQTL at the center of chromosome two.
        This vQTL acts additively on the log standard deviation scale, and results in residual standard deviation of [0.8, 1, 1.25] for the three genotype groups.
	\item \texttt{phenotype4} has an mvQTL at the center of chromosome three.
        This mvQTL has a mean effect that explains 2.7\% of phenotype variance and a variance effect that acts additively on the standard deviation scale, resulting in residual standard deviation of [0.85, 1, 1.17] for the three genotype groups.
\end{enumerate}

@@

The $p$-values described above, however, are nominal: they do not take into account multiple testing across the genome.
    They also rely on asymptotic theory that assumes the underlying phenotype being residually normal; this may not always be the case and when violated will lead to inflated significance.
More robust $p$-values that are corrected for genomewide significance via control of the family-wise error rate (FWER) can be obtained empirically, through a permutation procedure described below.

@@

Last, to assess the genetic precision of a discovered QTL for bioinformatic follow-up, the function \texttt{scanonevar.boot} estimates confidence intervals via the non-parametric bootstrap \citep{Visscher1996}.
This function takes, as arguments, a \texttt{scanonevar} object, the type of QTL detected, the name of the chromosome containing the QTL, and \texttt{num.resamples}, the number of bootstrap resamplings desired.
As with \texttt{scanonevar.perm}, the \texttt{n.cores} argument can be used to spread the bootstraps over many computational cores and defaults to the number of cores available minus two, and bootstraps can be run on separate computers and combined with \texttt{c} to increase the precision of the estimate of the confidence interval.

@@

For the first benchmark, we ran \texttt{scanonevar.perm} on the data from \citet{Kumar2013} and \CortyReanalysisPaper, which contains 244 individuals and 582 loci, varying the number of permutations desired and the number of computer cores used.
For a given number of cores, the relationship between time and the number of permutations is linear (\autoref{fig:benchmark_kumar}), the slope depending on the number of cores and ranging from $\approx$ 6.3 seconds per permutation with 4 cores to $\approx$ 1.2 second per permutation with 32 cores.

For the second benchmark, we ran \texttt{scanonevar.perm} on simulated data, always conducting 1000 permutations and using 32 cores, but varying the number of individuals in the mapping population and the number of markers in the genome.
For a given population size, there is a slightly curvilinear relationship between number of markers and time required (\autoref{fig:benchmark_sim_cross}), which reflects a linear increase in the time taken to conduct the permuted genome scans plus an increase in the time taken for ``bookkeeping'' tasks like organizing and reshaping genetic data.
The slope (minutes per locus) depends on the population size, ranging from $\approx$ 1.4 seconds per locus with a population of size 100 to $\approx$ 3.3 seconds per locus with a population of size 800.

Based on these benchmarks, the workflow presented here is practical for QTL mapping F2 intercross and similar populations on modern, multi-core scientific computers.
Populations with many recombinations, where dense genotyping arrays that interrogate $\textgreater$ 10,000 loci, could not be practically analyzed with package \texttt{vqtl} in this way, although it is likely that statistical and computational steps could be taken to make such studies feasible:
statistically, techniques could be used that allow for large-scale analysis without permutation testing \citep{Efron2004};
computationally, the software could be changed to run on a computer cluster, rather than on a single computer \citep{Jette2003a,Marchand2017}.

@@

\section{Acknowledgments}
This work was primarily funded by National Institutes of General Medical Sciences (NIGMS) grants to WV (R01-GM104125,R35-GM127000) and a National Institutes of Mental Health grant to RWC (F30-MH108265).
RWC received additional support from NIGMS grant T32-GM067553 and National Library of Medicine grant T32-LM012420.
